Supplementary Information

### 2021-02-04 writing/transect_paper.Rmd
RDP -> RDP-LSU

Assignment of taxonomic identities to environmental sequences is dependent on both the reference database and the algorithm used.
Although the public INSDC databases [@karsch-mizrachi2018] are often used for sequence identification, the open nature of submission to these databases results in a substantial fraction of incorrect taxonomic annotations [@nilsson2006;@steinegger2020] as well as sequences of poor technical quality [@nilsson2012].
<!--R4: Near L114 - More refs to consider: See Ashelford et al., 2005 Appl Env Microbiol 71: 7724,
Bidartondo et al., 2008 Science 319:1616, but also LeRay et al., 2019 PNAS 116: 22651 for a
more recent take-->
Consequently <!--R5: add comma--> taxonomic assignments based on these databases may be incorrect or inconsistent.
Several curated databases also exist which attempt to address these issues and which cover the whole fungal kingdom.
<!--R4: see Nilsson et al., 2005 BMC Bioinformtics 6: 178-->
The Unite database is a more curated <!--More curated?-->attempt to include all publically available ITS sequences (800\ 000 as of release 8.0), originally targetting fungi but now expanded to include all eukaryotes [@nilsson2019a], where efforts have been made to correct incorrect annotations and exclude low-quality sequences [@abarenkov2018a].
The Ribosomal Data Project [RDP, @cole2014] hosts two manually curated fungal barcode sequence databases, which are specifically intended for use in taxonomic assignment of sequences:
the Warcup ITS database, containing 18\ 000 manually curated fungal ITS sequences [@deshpande2016], and the RDP fungal LSU training set, containing 8000 manually curated LSU sequences from fungi and 3000 from other eukaryotic groups [@liu2012].
Although the quality of sequences and taxonomic annotations is undoubtedly higher in these more curated databases, they are inherently limited in taxonomic coverage and do not include the most recently published sequences.

@@

We compare three different sequencing platforms (PacBio RS II, Illumina MiSeq, Ion Torrent Ion S5), long and short amplicons, three different taxonomic assignment algorithms (RDP classifier, SINTAX, IDTAXA)
with three different reference databases (Unite, Warcup, RDP),
<!--R4: RDP reference sets include UNITE ITS and Warcup ITS as well as an LSU reference set
(Liu et al. 2012). I think it would be more clear to replace ...(Unite, Warcup, RDP) to (UNITE ITS,
Warcup ITS, Liu LSU)? As ‘RDP’ generally refers to the better-known prokaryote SSU rRNA gene
reference set.-->
and both non-phylogenetic and phylogenetic community distance measures.
We also present new algorithms for dividing the LSU into domains, combining denoising results from multiple domains as a strategy to capture more ASVs from long amplicons in diverse commmunities, and incorporating phylogenetic information into taxonomic assignments.

@@

Taxonomic annotations of the RDP LSU fungal training set version 11.5 [@cole2014] and Warcup ITS training set [@deshpande2016] were mapped to a uniform taxonomic classification system (see Supplementary Methods).
Preliminary taxonomic assignment was performed to genus level separately on the ITS region using Unite and Warcup and on the LSU region using RDP, respectively, as taxonomic references.
<!--R4: ...LSU region using RDP. I suggest you refer to the RDP LSU reference set (Liu et al., 2012)
instead of just to RDP which is better known for its prokaryote SSU rRNA gene classifier.-->
For each region/reference combination, taxonomy was assigned using three popular algorithms:
the RDP Naïve Bayesian Classifier [RDPC, @wang2007] as implemented in DADA2;
SINTAX [@edgar2016a] as implemented in VSEARCH v2.9.1 [@rognes2016];
and IDTAXA [@murali2018].
A relatively lax confidence threshhold of 50% was used for all three algorithms.
<!--R4: a relatively lax bootstrap support cutoff of 50% would indicate classification accuracy of
what? 75% accuracy, more, less, does it vary across ITS and LSU? I suggest this be addressed in
the main text, in https://rdp.cme.msu.edu/download/posters/fungalITSreport_062014.pdf see
Fig 3. -->
<!--R5: This seems very low. I think that some explanation for such a low value is needed to give a rationale. This is basically a coin toss so that is not exactly giving great confidence in the results-->
Each full-length ASV was thus given up to nine preliminary taxonomic assignments (three references $\times$ three algorithms).
ASVs from the short-amplicon datasets for which no matching long-amplicon ASV could be reconstructed were taxonomically assigned using Unite and Warcup on the full length of the short amplicon.

@@

(ref:taxon-chart) Fraction of ASVs (left) and reads (right) assigned to each taxonomic rank, for different sequencing technologies (PacBio RS II, Illumina MiSeq, Ion Torrent Ion S5), amplicons (Long, Short), reference databases (Unite, Warcup, RDP), and assignment algorithms (PHYLOTAX, Consensus, RDPC, SINTAX, IDTAXA).
Consensus and PHYLOTAX assignments are based on the consensus of RDPC, SINTAX, and IDTAXA, using all available databases and, in the case of PHYLOTAX, phylogenetic information.
<!--R4: This figure needs a ‘short’ title. Suggest you change RDP to RDP LSU training set. I’d
be interested in seeing a comparison of run times for each taxonomic assignment or a sentence
added to the main text comparing run times. In the discussion you may want to indicate that
phylogeny-based methods tend to be more computationally intensive compared to BLAST or
RDP ex. shown for fungal ITS (Porter and Golding, 2011 New Phyt 192: 775 ) and fungal LSU
(Porter and Golding, 2012 PLoS ONE 7: e35749).-->

@@

Assignment of ecological function to environmental fungal sequences is dependent on accurate taxonomic identification, especially at the genus level or below [@nguyen2016funguild].
However, different combinations of algortihms and reference datasets vary in their performance at confidently assigning taxonomy to sequences.
Although the RDP fungal training set and Unite performed comparably at taxonomic placement of long amplicon sequences, the Warcup database placed notably fewer sequences at all taxonomic levels for all datasets (Figure \@ref(fig:taxon-chart)).
This is probably due to two factors.
First, the Warcup database does not include any non-fungi, so it cannot place any non-fungal sequences.
Second, due to its low-density coverage of the fungal kingdom (18\ 000 sequences vs. 800\ 000 for Unite), it is likely that many ITS sequences, especially from uncultured tropical soil fungi, have no close match in the Warcup database, and so cannot be placed.
<!-- R4: replace ‘RDP fungal training set’ with RDP LSU training set (Liu et al., 2012) as the RDP
has 3 fungal training sets (UNITE ITS, Warcup ITS, LSU) in addition to the better-known
prokaryote training set SSU rRNA.-->
The RDP fungal training set, which has even fewer sequences (8000 fungi plus 3000 other eukaryotes), is probably more successful due to higher sequence conservation in LSU.
@heeger2019 also found that a more conserved region, in their case 5.8S, outperformed ITS at placing sequences without close database matches.
Of the three algorithms tested, IDTAXA placed fewer sequences than RDPC or SINTAX with all databases, as expected given its more well-calibrated and conservative confidence scores [@murali2018a], but this was particularly dramatic when paired with the Warcup database, where IDTAXA placed <25% of ASVs even to phylum.
<!--R5: This discussion of placement is good. However, it suffers from one serious weakness. It seems to assume that when things are placed that they are placed CORRECTLY. In my experience some things are obviously incorrectly placed. You should explicitly say whether you actually dealt with his problem or not. If not, at least give some caveats here!-->

@@

 - Trimmed, demultiplexed sequencing reads have been deposited at the European Nucleotide Archive (ERA) under Project accession number PRJEB37385. Accession numbers are given in Supplementary Files 1 and 2.
 - Consensus ASV sequences will also be deposited at ENA prior to publication.
 - Nucleotide alignment and ML tree will be deposited at Dryad prior to final publication [@furneauxdata2020].
 - R packages `LSUx`, `tzara`, `phylotax`, and `FUNGuildR` are available on Github at https://github.com/brendanf/LSUx, https://github.com/brendanf/tzara, https://github.com/brendanf/phylotax, and https://github.com/brendanf/FUNGuildR.
These packages are currently being prepared for submission to CRAN/Bioconductor.
If they are not accepted prior to final publication, snapshots will be archived at Dryad.
 - FASTA-format files for the RDP, Warcup, and Unite reference databases with unified classifications, as well as scripts used to generate them, are available at https://github.com/brendanf/reannotate. The versions used in this paper will be archived at Dryad prior to publication.
 - Bioinformatics pipeline and analysis scripts are available at https://github.com/oueme-fungi/oueme-fungi-transect.

>>>>

Assignment of taxonomic identities to environmental sequences is dependent on both the reference database and the algorithm used.
Although the public INSDC databases [@karsch-mizrachi2018] are often used for sequence identification, the open nature of submission to these databases results in a substantial fraction of incorrect taxonomic annotations [@nilsson2006;@steinegger2020] as well as sequences of poor technical quality [@nilsson2012].
<!--R4: Near L114 - More refs to consider: See Ashelford et al., 2005 Appl Env Microbiol 71: 7724,
Bidartondo et al., 2008 Science 319:1616, but also LeRay et al., 2019 PNAS 116: 22651 for a
more recent take-->
Consequently <!--R5: add comma--> taxonomic assignments based on these databases may be incorrect or inconsistent.
Several curated databases also exist which attempt to address these issues and which cover the whole fungal kingdom.
<!--R4: see Nilsson et al., 2005 BMC Bioinformtics 6: 178-->
The Unite database is a more curated <!--More curated?-->attempt to include all publically available ITS sequences (800\ 000 as of release 8.0), originally targetting fungi but now expanded to include all eukaryotes [@nilsson2019a], where efforts have been made to correct incorrect annotations and exclude low-quality sequences [@abarenkov2018a].
The Ribosomal Data Project [RDP, @cole2014] hosts two manually curated fungal barcode sequence databases, which are specifically intended for use in taxonomic assignment of sequences:
the Warcup ITS training set, containing 18\ 000 manually curated fungal ITS sequences [@deshpande2016], and the RDP fungal LSU training set, containing 8000 manually curated LSU sequences from fungi and 3000 from other eukaryotic groups [RDP-LSU, @liu2012].
Although the quality of sequences and taxonomic annotations is undoubtedly higher in these more curated databases, they are inherently limited in taxonomic coverage and do not include the most recently published sequences.

@@

We compare three different sequencing platforms (PacBio RS II, Illumina MiSeq, Ion Torrent Ion S5), long and short amplicons, three different taxonomic assignment algorithms (RDPC, SINTAX, IDTAXA)
with three different reference databases (Unite, Warcup, RDP-LSU),
and both non-phylogenetic and phylogenetic community distance measures.
We also present new algorithms for dividing the LSU into domains, combining denoising results from multiple domains as a strategy to capture more ASVs from long amplicons in diverse commmunities, and incorporating phylogenetic information into taxonomic assignments.

@@

Taxonomic annotations of the RDP-LSU training set version 11.5 [@liu2012;@cole2014] and Warcup ITS training set [@deshpande2016] were mapped to a uniform taxonomic classification system (see Supplementary Methods).
Preliminary taxonomic assignment was performed to genus level separately on the ITS region using Unite and Warcup and on the LSU region using RDP-LSU, respectively, as taxonomic references.
For each region/reference combination, taxonomy was assigned using three popular algorithms:
the RDPC [@wang2007] as implemented in DADA2;
SINTAX [@edgar2016a] as implemented in VSEARCH v2.9.1 [@rognes2016];
and IDTAXA [@murali2018].
A relatively lax confidence threshhold of 50% was used for all three algorithms.
<!--R4: a relatively lax bootstrap support cutoff of 50% would indicate classification accuracy of
what? 75% accuracy, more, less, does it vary across ITS and LSU? I suggest this be addressed in
the main text, in https://rdp.cme.msu.edu/download/posters/fungalITSreport_062014.pdf see
Fig 3. -->
<!--R5: This seems very low. I think that some explanation for such a low value is needed to give a rationale. This is basically a coin toss so that is not exactly giving great confidence in the results-->
Each full-length ASV was thus given up to nine preliminary taxonomic assignments (three references $\times$ three algorithms).
ASVs from the short-amplicon datasets for which no matching long-amplicon ASV could be reconstructed were taxonomically assigned using Unite and Warcup on the full length of the short amplicon.

@@

(ref:taxon-chart) Fraction of ASVs (left) and reads (right) assigned to each taxonomic rank, for different sequencing technologies (PacBio RS II, Illumina MiSeq, Ion Torrent Ion S5), amplicons (Long, Short), reference databases (Unite, Warcup, RDP-LSU), and assignment algorithms (PHYLOTAX, Consensus, RDPC, SINTAX, IDTAXA).
Consensus and PHYLOTAX assignments are based on the consensus of RDPC, SINTAX, and IDTAXA, using all available databases and, in the case of PHYLOTAX, phylogenetic information.
<!--R4: This figure needs a ‘short’ title. I’d
be interested in seeing a comparison of run times for each taxonomic assignment or a sentence
added to the main text comparing run times. In the discussion you may want to indicate that
phylogeny-based methods tend to be more computationally intensive compared to BLAST or
RDP ex. shown for fungal ITS (Porter and Golding, 2011 New Phyt 192: 775 ) and fungal LSU
(Porter and Golding, 2012 PLoS ONE 7: e35749).-->

@@

Assignment of ecological function to environmental fungal sequences is dependent on accurate taxonomic identification, especially at the genus level or below [@nguyen2016funguild].
However, different combinations of algortihms and reference datasets vary in their performance at confidently assigning taxonomy to sequences.
Although RDP-LSU and Unite performed comparably at taxonomic placement of long amplicon sequences, the Warcup database placed notably fewer sequences at all taxonomic levels for all datasets (Figure \@ref(fig:taxon-chart)).
This is probably due to two factors.
First, the Warcup database does not include any non-fungi, so it cannot place any non-fungal sequences.
Second, due to its low-density coverage of the fungal kingdom (18\ 000 sequences vs. 800\ 000 for Unite), it is likely that many ITS sequences, especially from uncultured tropical soil fungi, have no close match in the Warcup database, and so cannot be placed.
RDP-LSU, which has even fewer sequences (8000 fungi plus 3000 other eukaryotes), is probably more successful due to higher sequence conservation in LSU.
@heeger2019 also found that a more conserved region, in their case 5.8S, outperformed ITS at placing sequences without close database matches.
Of the three algorithms tested, IDTAXA placed fewer sequences than RDPC or SINTAX with all databases, as expected given its more well-calibrated and conservative confidence scores [@murali2018a], but this was particularly dramatic when paired with the Warcup database, where IDTAXA placed <25% of ASVs even to phylum.
<!--R5: This discussion of placement is good. However, it suffers from one serious weakness. It seems to assume that when things are placed that they are placed CORRECTLY. In my experience some things are obviously incorrectly placed. You should explicitly say whether you actually dealt with his problem or not. If not, at least give some caveats here!-->

@@

 - Trimmed, demultiplexed sequencing reads have been deposited at the European Nucleotide Archive (ERA) under Project accession number PRJEB37385. Accession numbers are given in Supplementary Files 1 and 2.
 - Consensus ASV sequences will also be deposited at ENA prior to publication.
 - Nucleotide alignment and ML tree will be deposited at Dryad prior to final publication [@furneauxdata2020].
 - R packages `LSUx`, `tzara`, `phylotax`, and `FUNGuildR` are available on Github at https://github.com/brendanf/LSUx, https://github.com/brendanf/tzara, https://github.com/brendanf/phylotax, and https://github.com/brendanf/FUNGuildR.
These packages are currently being prepared for submission to CRAN/Bioconductor.
If they are not accepted prior to final publication, snapshots will be archived at Dryad.
 - FASTA-format files for the RDP fungal LSU training set, Warcup, and Unite reference databases with unified classifications, as well as scripts used to generate them, are available at https://github.com/brendanf/reannotate. The versions used in this paper will be archived at Dryad prior to publication.
 - Bioinformatics pipeline and analysis scripts are available at https://github.com/oueme-fungi/oueme-fungi-transect.

### 2021-02-04 writing/transect_supplement.Rmd
add line for rarefaction in accumulation plots
RDP -> RDP-LSU

**Supplementary file 3**: ML tree of all long metabarcoding ASVs (ITS1--5.8S--ITS2--LSU) assigned to kingdom Fungi.
Thick branches represent rapid bootstrap values >90%.
Taxonomic assignment by PHYLOTAX, based on assignments using RDPC, SINTAX, and IDTAXA, with Unite, Warcup, and RDP reference databases. 
Taxon names in blue were not fully resolved with PHYLOTAX, and are given as a list of alternative assignments, along with the number of methods which gave each assignment (max 9).
Taxon names in red were reconstructed as polyphyletic.
In some cases, this may be due to a poorly identified sequence "splitting" the group (e.g., the single *Monascus* sequence nested in *Aspergillus*; Aspergillaceae and Trichocomaceae sequences nested in Onygenales), but in other cases, the tree has clearly failed to reconstruct groups which are generally considered to be monophyletic based on multi-gene trees (e.g., Dothidiomycetes, Cantharellales).

@@

### Uniform taxonomic classification for Unite, Warcup, and RDP

The taxonomic classification system is based on the Unite database version 8 [@nilsson2019a].
In particular, the classification for fungi was according to @tedersoo2018, and for non-fungal eukaryotes was according to the proposed system of @tedersoo2017c as described in [@tedersoodata2017].
Although the latter system is not formally published, it is consistent with the annotations for non-fungal eukaryotes in the Unite database.
Additionally, it is a system with both purportedly monophyletic taxa and a uniform set of taxon ranks, which make it more appropriate for sequence-based taxonomic assignment algorithms than more accepted classification systems such as that of the International Society of Protistologists [@adl2019], which utilizes hierarchical nameless ranks.
FASTA-format files for the RDP, Warcup, and Unite reference databases with unified classifications, as well as scripts used to generate them, are available at https://github.com/brendanf/reannotate.

@@

(ref:accum-OTU) Each curve represents rarefaction of a single sample or pooled pair of samples. Points at the end of each curve represent the actual read depth and observed OTU richness. Panel at left has enlarged scale to show PacBio more clearly.

@@

(ref:taxa) Color and size of nodes represent the fractional ASV richness of that taxonomic group.
Color and thickness of branches represent the fractional read abundance of all ASVs belonging to that taxonomic group.
Diverse but relatively rare groups are thus shown as large, dark-colored nodes on relatively thin, light-colored branches (e.g., Dothideomycetes), while common but relatively non-diverse groups are shown as small, light-colored nodes on relatively thick, dark-colored branches (e.g., Russulales).
Groups which are not represented by at least 1% of reads or 1% of ASVs in any dataset are collapsed into nodes labeled "*".
ASVs which could not be further assigned are labelled "?".
ASV richness and read abundance are displayed as the mean across sequencing runs.
Taxonomic assignment is by strict consensus of multiple ITS databases (Unite, Warcup) and assignment algorithms (RDPC, SINTAX, IDTAXA).
The RDP LSU database was not included, to ensure consistent assignment between long and short amplicon libraries.

@@

(ref:ecm) Reads which were not identified to kingdom, or which were identified as a non-Fungi kingdom, are not included.
"unidentified" denotes ASVs which were assigned to kingdom Fungi, but could not be identified to family level or could not be assigned ECM status based on the available identification.
Height of each bar represents the fraction of reads represented by that bar.
Results are shown for different sequencing technologies, amplicons (Long, Short), reference databases (Unite, Warcup, RDP; "All" denotes algorithms which combine results from multiple databases), and taxonomic assignment algorithms.

@@

(ref:ecm-heattree) Red-orange edges (nodes) represent increased read (ASV) count for the short amplicon library, while blue-green edges (nodes) represent increased read (ASV) count for the long amplicon library.
Values are log~10~ of the ratio of mean values across all sequencing technologies for each amplicon.
Groups which are not represented by at least 1% of reads or 1% of ASVs in any dataset are collapsed into nodes labeled "*".
ASVs which could not be further assigned are labelled "?".
Taxonomic assignment is by strict consensus of multiple ITS databases (Unite, Warcup) and assignment algorithms (RDPC, SINTAX, IDTAXA).
The RDP LSU database was not included, to ensure consistent assignment between long and short amplicon libraries.

>>>>

**Supplementary file 3**: ML tree of all long metabarcoding ASVs (ITS1--5.8S--ITS2--LSU) assigned to kingdom Fungi.
Thick branches represent rapid bootstrap values >90%.
Taxonomic assignment by PHYLOTAX, based on assignments using RDPC, SINTAX, and IDTAXA, with Unite, Warcup, and RDP-LSU reference databases. 
Taxon names in blue were not fully resolved with PHYLOTAX, and are given as a list of alternative assignments, along with the number of methods which gave each assignment (max 9).
Taxon names in red were reconstructed as polyphyletic.
In some cases, this may be due to a poorly identified sequence "splitting" the group (e.g., the single *Monascus* sequence nested in *Aspergillus*; Aspergillaceae and Trichocomaceae sequences nested in Onygenales), but in other cases, the tree has clearly failed to reconstruct groups which are generally considered to be monophyletic based on multi-gene trees (e.g., Dothidiomycetes, Cantharellales).

@@

### Uniform taxonomic classification for Unite, Warcup, and RDP-LSU

The taxonomic classification system is based on the Unite database version 8 [@nilsson2019a].
In particular, the classification for fungi was according to @tedersoo2018, and for non-fungal eukaryotes was according to the proposed system of @tedersoo2017c as described in [@tedersoodata2017].
Although the latter system is not formally published, it is consistent with the annotations for non-fungal eukaryotes in the Unite database.
Additionally, it is a system with both purportedly monophyletic taxa and a uniform set of taxon ranks, which make it more appropriate for sequence-based taxonomic assignment algorithms than more accepted classification systems such as that of the International Society of Protistologists [@adl2019], which utilizes hierarchical nameless ranks.
FASTA-format files for the RDP-LSU, Warcup, and Unite reference databases with unified classifications, as well as scripts used to generate them, are available at https://github.com/brendanf/reannotate.

@@

(ref:accum-OTU) Each curve represents rarefaction of a single sample or pooled pair of samples.
Points at the end of each curve represent the actual read depth and observed OTU richness.
Panel at left has enlarged scale to show PacBio more clearly.
Vertical dashed line at 100 reads indicates rarefaction level for OTU richness comparisons.

@@

(ref:taxa) Color and size of nodes represent the fractional ASV richness of that taxonomic group.
Color and thickness of branches represent the fractional read abundance of all ASVs belonging to that taxonomic group.
Diverse but relatively rare groups are thus shown as large, dark-colored nodes on relatively thin, light-colored branches (e.g., Dothideomycetes), while common but relatively non-diverse groups are shown as small, light-colored nodes on relatively thick, dark-colored branches (e.g., Russulales).
Groups which are not represented by at least 1% of reads or 1% of ASVs in any dataset are collapsed into nodes labeled "*".
ASVs which could not be further assigned are labelled "?".
ASV richness and read abundance are displayed as the mean across sequencing runs.
Taxonomic assignment is by strict consensus of multiple ITS databases (Unite, Warcup) and assignment algorithms (RDPC, SINTAX, IDTAXA).
The RDP-LSU database was not included, to ensure consistent assignment between long and short amplicon libraries.

@@

(ref:ecm) Reads which were not identified to kingdom, or which were identified as a non-Fungi kingdom, are not included.
"unidentified" denotes ASVs which were assigned to kingdom Fungi, but could not be identified to family level or could not be assigned ECM status based on the available identification.
Height of each bar represents the fraction of reads represented by that bar.
Results are shown for different sequencing technologies, amplicons (Long, Short), reference databases (Unite, Warcup, RDP-LSU; "All" denotes algorithms which combine results from multiple databases), and taxonomic assignment algorithms.

@@

(ref:ecm-heattree) Red-orange edges (nodes) represent increased read (ASV) count for the short amplicon library, while blue-green edges (nodes) represent increased read (ASV) count for the long amplicon library.
Values are log~10~ of the ratio of mean values across all sequencing technologies for each amplicon.
Groups which are not represented by at least 1% of reads or 1% of ASVs in any dataset are collapsed into nodes labeled "*".
ASVs which could not be further assigned are labelled "?".
Taxonomic assignment is by strict consensus of multiple ITS databases (Unite, Warcup) and assignment algorithms (RDPC, SINTAX, IDTAXA).
The RDP-LSU database was not included, to ensure consistent assignment between long and short amplicon libraries.
